---
title: 'Practical Machine Learning: Course Project'
author: "Alejandro Osorio"
date: "August, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Getting and Cleaning Data

Training and testing sets were downloaded using read_csv function, with the 'na' parameter tweaked so that it also included "#DIV/0!" cases (besides the default "" and "NA" cases).  Additionally, first columns containing row numbers were eliminated from the obtained datasets.

```{r, include=FALSE}
training <- read_csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", na = c("", "NA", "#DIV/0!"))
training <- training[,-1]
testing <- read_csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", na = c("", "NA"))
testing <- testing[,-1]
```

Next, columns from the testing set that included NAs, where identified and dimensioned the following way:

```{r}
table(colSums((is.na(testing))))
```

Therefore, as all columns with NAs (100 in this case) had 100% NAs (20 each), those columns were removed both from the testing and training datasets.

```{r, echo=FALSE}
datacols <- names(which(colSums(is.na(testing))!=0))
testing <- select(testing, -datacols)
training <- select(training, -datacols)
```

Finally, the training dataset with which to start working, went down to the following characteristics:

```{r}
str(training)
```


## Including Plots

You can also embed plots, for example:
